% =============================================================================
% IsalSR Technical Report
% =============================================================================
% Authors: Ezequiel Lopez-Rubio, Mario Pascual Gonzalez
% University of Malaga, 2025-2026
%
% PURPOSE: Serves as basis for the arXiv preprint and IEEE TPAMI submission.
% Provides complete context for the PhD advisor (Ezequiel) regarding
% what was implemented, how it works, and what experiments validate.
%
% STRUCTURE: Each \input loads a section file. Comments above each \input
% describe its content so agents can use main.tex as a navigation hub.
% =============================================================================

\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{multirow}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

\title{IsalSR: Instruction Set and Language for Symbolic Regression\\[6pt]
\large Technical Report --- Implementation and Experimental Design}

\author{
  Ezequiel Lopez-Rubio\textsuperscript{1} \and
  Mario Pascual Gonzalez\textsuperscript{1} \\[6pt]
  \textsuperscript{1}University of Malaga, Spain \\
  \texttt{\{ezeqlr, mpascual\}@uma.es}
}

\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
We present IsalSR (Instruction Set and Language for Symbolic Regression),
a representation framework that encodes mathematical expressions as labeled
Directed Acyclic Graphs (DAGs) via isomorphism-invariant instruction strings.
By computing a canonical string---a complete labeled-DAG invariant---IsalSR
collapses $O(k!)$ equivalent expression representations into a single
canonical form, where $k$ is the number of internal (non-variable) nodes.
This reduction enables existing symbolic regression methods to search over a
dramatically smaller space without modifying their core algorithms.
This report describes the instruction set design, canonical string computation,
software architecture, and experimental infrastructure for validating the
$O(k!)$ claim on standard SR benchmarks (Nguyen, Feynman, SRBench).
The implementation comprises 286 verified tests across 7 development phases,
with 78/80 proposal-alignment checks passed.
\end{abstract}

% --- Section 1: Proposal and positioning ---
% WHERE IsalSR fits in the SR landscape, HOW we orient the narrative around
% search space reduction, and WHY this is a publishable contribution.
\input{sections/01_proposal}

% --- Section 2: Instruction set and core structures ---
% The IsalSR alphabet Sigma_SR (two-tier encoding), LabeledDAG, CDLL,
% initial state convention, edge semantics, and the DAG constraint.
\input{sections/02_instruction_set}

% --- Section 3: Core algorithms ---
% StringToDAG (S2D decoder), DAGToString (D2S greedy encoder), canonical
% string (exhaustive + pruned with 6-component structural tuples),
% round-trip property, and Levenshtein DAG distance.
\input{sections/03_algorithms}

% --- Section 4: Software architecture ---
% Dependency layering (core zero-deps -> adapters -> evaluation -> search),
% module inventory, testing strategy (286 tests), code quality standards.
\input{sections/04_architecture}

% --- Section 5: Benchmarks and experimental design ---
% Exact Nguyen (12) and Feynman (10) benchmarks from Liu2025, metrics
% (R^2, NRMSE, reward), WITH vs WITHOUT canonicalization comparison design.
\input{sections/05_benchmarks}

% --- Section 6: Preliminary results ---
% Local search space analysis demonstrating empirical O(k!) reduction,
% and sample search results on Nguyen benchmarks.
\input{sections/06_results}

% --- Section 8: Precomputed cache ---
% Data structure, storage format, and deployment for downstream analysis.
\input{sections/08_precomputed_cache}

% --- Section 9: Conclusion and next steps ---
% Summary of contributions, limitations, and roadmap to IEEE TPAMI submission.
\input{sections/07_conclusion}

\begin{thebibliography}{10}

\bibitem{lopezrubio2025}
E.~Lopez-Rubio, ``Representation of the structure of graphs by sequences of
instructions,'' \emph{arXiv:2512.10429v2}, 2025.

\bibitem{liu2025}
J.~Liu et~al., ``Mathematical expression exploration with graph representation
and generative graph neural network,'' \emph{Neural Networks}, vol.~187,
p.~107405, 2025.

\bibitem{lacava2021}
W.~La~Cava et~al., ``Contemporary symbolic regression methods and their
relative performance,'' in \emph{NeurIPS Datasets and Benchmarks}, 2021.

\bibitem{petersen2021}
B.~K. Petersen et~al., ``Deep symbolic regression: Recovering mathematical
expressions from data via risk-seeking policy gradients,'' in \emph{ICLR}, 2021.

\bibitem{uy2011}
N.~Q. Uy et~al., ``Semantically-based crossover in genetic programming:
Application to real-valued symbolic regression,'' \emph{Genetic Programming
and Evolvable Machines}, vol.~12, pp.~91--119, 2011.

\bibitem{udrescu2020}
S.-M. Udrescu and M.~Tegmark, ``AI Feynman: A physics-inspired method for
symbolic regression,'' \emph{Science Advances}, vol.~6, no.~16, 2020.

\bibitem{koza1992}
J.~R. Koza, \emph{Genetic Programming: On the Programming of Computers by
Means of Natural Selection}.\hskip 1em MIT Press, 1992.

\bibitem{wilson1927}
E.~B. Wilson, ``Probable inference, the law of succession, and statistical
inference,'' \emph{Journal of the American Statistical Association}, vol.~22,
no.~158, pp.~209--212, 1927.

\end{thebibliography}

\end{document}
